---
title: "TMA4268 - Project 1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 2

### a)

One error made is excluding sex because of a low p-value, the p-value is the probability to observe a result equal or more extreme than the one we did, given that the null hypothesis $H_0: \beta = 0$, so a lower p-value means the result is more statistically significant.

Another error made is deciding whether to reject the null hypothesis that the species coefficient overall is actually zero based on the p-values of the coefficients. Since the null hypothesis in this case is $H_0: \beta_{Chinstrap} = \beta_{Gentoo} = 0$ at the same time. We need to do a F-test.


### b)

```{r}
library(ggplot2)
library(palmerpenguins) # Contains the data set 'penguins'.
data(penguins)

Penguins <- subset(penguins, select = -c(island, year))
Penguins <- na.omit(Penguins)

ggplot(data = Penguins) +
  geom_point(mapping = aes(x = bill_depth_mm, y = body_mass_g, colour = species))

ggplot(data = Penguins) +
  geom_histogram(binwidth = 40, mapping = aes(x = body_mass_g)) +
  facet_wrap(~ sex, nrow = 2)

```

### c)

```{r}
Penguins <- subset(penguins, select = -c(island, year))

penguin.model <- lm(body_mass_g ~ flipper_length_mm + sex + bill_depth_mm * species,
                    data = Penguins)
summary(penguin.model)

anova(penguin.model)




```




